\documentclass[12pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\pgfplotsset{compat=1.18}
\geometry{margin=1in}

% Theorems and definitions
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}

% Title
\title{Honjo Masamune: Technical Specification of a Biomimetic Metacognitive Truth Engine}
\author{ }
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This document specifies the Honjo Masamune system: a biomimetic metacognitive engine that reconstructs consistent world-states from incomplete, decaying, and adversarially perturbed evidence streams. The system integrates (i) a temporal Bayesian learning engine with resource-aware variational objectives, (ii) an adversarial hardening subsystem formalized as a minimax optimization over evidence transformations and graph topologies, (iii) a decision optimization engine defined as a cost-aware Markov Decision Process (MDP) augmented with stochastic dynamics, and (iv) an orchestration module that selects among ensemble integration patterns as a function of query complexity and computational budget. All objectives are explicitly regularized by a resource cost functional that models computational metabolism. Algorithms, equations, and diagrams are provided to permit independent reimplementation.
\end{abstract}

\tableofcontents

\section{System Overview}
\subsection{Notation}
We denote by $\mathcal{E} = \{e_i\}_{i=1}^{N}$ a set of evidence items. Each evidence item has features $x_i \in \mathbb{R}^{d}$, metadata $m_i$, timestamp $t_i$, and a source descriptor $s_i$. The latent world state is $\mathbf{z} \in \mathcal{Z}$; derived claims are $\mathbf{y} \in \mathcal{Y}$. A belief network is represented as a directed acyclic or loopy graph $\mathcal{G}=(\mathcal{V},\mathcal{A})$ with node latents $\{\mathbf{z}_v\}_{v \in \mathcal{V}}$. Resource ("metabolic") costs are modeled by a nonnegative functional $\mathcal{C}(\cdot)$ with units analogous to ATP.

\subsection{Core Modules}
Honjo Masamune consists of four modules:
\begin{enumerate}[label=\textbf{M\arabic*}., leftmargin=*]
  \item Temporal Bayesian learning engine (Mzekezeke): evidence assimilation and posterior maintenance with temporal decay.
  \item Adversarial hardening system (Diggiden): continuous robustness optimization against structured perturbations.
  \item Decision optimization engine (Hatata): resource- and risk-aware control for state transitions.
  \item Orchestration and ensemble integration (Diadochi): complexity-aware expert combination with cost-sensitive routing.
\end{enumerate}

\section{Temporal Bayesian Learning (Mzekezeke)}
\subsection{Evidence Decay and Weighting}
Evidence utility degrades with time and context. For an evidence item $e_i$ observed at time $t_i$, define a family of decay functions
\begin{equation}
\label{eq:decay}
\omega_i(\Delta t; \bm{\phi}) \in (0,1], \quad \Delta t = t - t_i,
\end{equation}
where $\bm{\phi}$ parameterizes a selected decay model. The system supports, e.g.,
\begin{align}
\text{Exponential:}\quad & \omega_i = \exp(-\lambda_i \Delta t), \\
\text{Power law:}\quad & \omega_i = (1 + \kappa_i \Delta t)^{-\alpha_i}, \\
\text{Logistic:}\quad & \omega_i = \bigl(1 + \exp(\beta_i(\Delta t - \tau_i))\bigr)^{-1}.
\end{align}
Weights modulate both the likelihood and the contribution to resource costs.

\subsection{Generative Model and Variational Objective}
Let the generative model factorize as
\begin{equation}
\label{eq:generative}
 p(\mathbf{x}, \mathbf{y}, \mathbf{z} \mid \Theta) = p(\mathbf{z}) \; p(\mathbf{y}\mid \mathbf{z}) \; \prod_{i=1}^{N} p\bigl(x_i \mid \mathbf{z}, m_i, s_i\bigr)^{\omega_i},
\end{equation}
where $\Theta$ denotes global parameters. The variational posterior $q_{\Phi}(\mathbf{z})$ is obtained by minimizing a resource-regularized negative Evidence Lower Bound (ELBO):
\begin{equation}
\label{eq:relbo}
\mathcal{J}_{\text{learn}}(\Phi,\Theta) = \underbrace{\mathrm{KL}\bigl(q_{\Phi}(\mathbf{z})\,\Vert\, p(\mathbf{z})\bigr) - \mathbb{E}_{q_{\Phi}}\bigl[\log p(\mathbf{y}\mid \mathbf{z})\bigr] - \sum_{i=1}^N \omega_i\, \mathbb{E}_{q_{\Phi}}\bigl[\log p(x_i \mid \mathbf{z}, m_i, s_i)\bigr]}_{\text{negative ELBO}} 
+ \; \lambda_{\text{ATP}} \, \mathcal{C}_{\text{learn}}(\Phi,\Theta),
\end{equation}
where $\lambda_{\text{ATP}}>0$ trades off fit versus cost. The cost functional admits a decomposition
\begin{equation}
\label{eq:cost}
\mathcal{C}_{\text{learn}} = c_{\text{inference}} + c_{\text{storage}} + c_{\text{io}} + c_{\text{verification}},
\end{equation}
with each term measurable (e.g., flops, memory, latency) and convertible to a unified unit via coefficients calibrated to a hardware baseline.

\subsection{Graphical Structure and Temporal Smoothing}
When $\mathcal{G}$ is dynamic, nodes follow a smoothing prior:
\begin{equation}
\label{eq:rw}
 p(\mathbf{z}_v^{(t)} \mid \mathbf{z}_v^{(t-1)}) = \mathcal{N}\bigl(\mathbf{z}_v^{(t)}; \mathbf{z}_v^{(t-1)}, \Sigma_v\bigr),\qquad \Sigma_v \succ 0,
\end{equation}
with cross-node couplings through sparse potentials $\psi_{uv}(\mathbf{z}_u, \mathbf{z}_v)$. The inference employs amortized variational families respecting graph sparsity to maintain linear- or near-linear-time updates in the number of edges.

\section{Adversarial Hardening (Diggiden)}
\subsection{Threat Model}
Define a family of admissible attacks $\mathcal{A}$ acting on evidence, metadata, timestamps, and topology:
\begin{equation}
\label{eq:attacks}
 a: (\{x_i,m_i,t_i,s_i\}, \mathcal{G}) \mapsto (\{x'_i,m'_i,t'_i,s'_i\}, \mathcal{G}'), \quad a \in \mathcal{A}.
\end{equation}
Attacks include contradictory injection, timestamp skew, source spoofing, edge rewiring, and feature perturbations bounded in normed or f-divergence balls.

\subsection{Robust Optimization}
Robust learning solves
\begin{equation}
\label{eq:minimax}
 \min_{\Phi,\Theta} \; \max_{a \in \mathcal{A}} \; \mathcal{J}_{\text{learn}}\bigl(\Phi,\Theta; a(\mathcal{E},\mathcal{G})\bigr) + \lambda_{\text{ATP}}\, \mathcal{C}_{\text{adv}}(a),
\end{equation}
with $\mathcal{C}_{\text{adv}}(a)$ penalizing expensive attacks during training to emphasize realistic threat surfaces. Attack generation uses a portfolio of strategies sampled from a policy $\pi_{\text{adv}}$ updated by success-weighted reinforcement to target current weaknesses.

\subsection{Property-Based Fuzzing for Structural Coverage}
Let $\mathcal{P}$ be a set of invariants (e.g., acyclicity constraints, calibration bounds). Fuzzers sample transformations $a\sim \Pi$ to maximize violation scores $V(a;\mathcal{P})$ under cost constraints, driving test coverage of rare graph states while maintaining bounded $\mathcal{C}_{\text{adv}}$.

\section{Decision Optimization (Hatata)}
\subsection{MDP with Resource-Aware Rewards}
Define an MDP $(\mathcal{S},\mathcal{A},P,r,\gamma)$ where states summarize posteriors and system context; actions trigger computation and integration steps; transitions $P$ follow execution results. Rewards are cost-adjusted utilities
\begin{equation}
\label{eq:reward}
 r(s,a) = U\bigl(\text{consistency}(s),\, \text{calibration}(s),\, \text{robustness}(s)\bigr) - \eta\, \mathcal{C}_{\text{exec}}(s,a), \quad \eta>0.
\end{equation}
Policies $\pi$ are obtained via value iteration or policy optimization; for continuous-time or continuous-state dynamics, we augment with stochastic differential control.

\subsection{Stochastic Dynamics Augmentation}
For selected summaries $X_t$, we model
\begin{equation}
\label{eq:sde}
 dX_t = \mu(X_t, a_t)\,dt + \sigma(X_t,a_t)\, dW_t + dJ_t,
\end{equation}
with $W_t$ a Wiener process and $J_t$ optional jump terms. Control selects $a_t$ to optimize discounted returns under risk-sensitive criteria (e.g., entropic risk).

\section{Orchestration and Expert Integration (Diadochi)}
\subsection{Complexity-Conditioned Pattern Selection}
Let $c \in [0,1]$ denote query complexity (estimated from ambiguity, heterogeneity, and evidence conflict) and $B$ a budget. Define a discrete selector
\begin{equation}
\label{eq:selector}
 \kappa(c) = \begin{cases}
  \text{router} & c \in [0, c_1),\\
  \text{prompt-composition} & c \in [c_1, c_2),\\
  \text{sequential-chain} & c \in [c_2, c_3),\\
  \text{mixture-of-experts} & c \in [c_3, 1],
 \end{cases}
\end{equation}
subject to $\mathcal{C}_{\text{pattern}}(\kappa(c)) \le B$.

\subsection{Mixture-of-Experts (MoE)}
Given expert outputs $\{h_k\}_{k=1}^{K}$, a gating network produces weights
\begin{equation}
\label{eq:moe}
 w_k = \frac{\exp(g_k(\xi))}{\sum_{j=1}^{K} \exp(g_j(\xi))}, \quad \hat{y} = \sum_{k=1}^{K} w_k h_k,
\end{equation}
where $\xi$ summarizes the query and posterior diagnostics. Alternative weightings include linear, binary selection, or learned convex combinations with sparsity regularization. Pattern selection is jointly optimized with $\pi$ in Hatata.

\section{Resource Cost Model (Computational Metabolism)}
\subsection{Unified Cost Functional}
All modules contribute to a unified cost
\begin{equation}
\label{eq:totalcost}
 \mathcal{C}_{\text{total}} = \mathcal{C}_{\text{learn}} + \mathcal{C}_{\text{adv}} + \mathcal{C}_{\text{control}} + \mathcal{C}_{\text{orchestration}},
\end{equation}
measured by calibrated surrogates of energy, time, and memory. Objectives incorporate $\mathcal{C}_{\text{total}}$ via Lagrange multipliers to enforce budgets.

\section{Algorithms}
\subsection{End-to-End Cycle}
\begin{algorithm}[H]
\caption{Honjo Masamune Cycle}
\label{alg:cycle}
\begin{algorithmic}[1]
\State \textbf{Input:} Evidence $\mathcal{E}$, graph $\mathcal{G}$, budget $B$, complexity estimate $c$, hyperparameters.
\State Select integration pattern $p \gets \kappa(c)$ with $\mathcal{C}_{\text{pattern}}(p) \le B$.
\State Update variational posterior by minimizing \eqref{eq:relbo} with decay weights \eqref{eq:decay}.
\For{$t=1\ldots T_{\text{adv}}$}
  \State Sample attack $a_t \sim \pi_{\text{adv}}$; compute robust gradient of \eqref{eq:minimax}.
  \State Update $(\Phi,\Theta)$ and adversary policy parameters.
\EndFor
\State Solve/approximate control policy $\pi$ for \eqref{eq:reward} under dynamics \eqref{eq:sde}.
\State Produce integrated reconstruction $\hat{y}$ via pattern $p$ and MoE \eqref{eq:moe} as applicable.
\State \textbf{Output:} Posteriors, reconstruction $\hat{y}$, diagnostics, cost ledger.
\end{algorithmic}
\end{algorithm}

\subsection{Adversarial Portfolio Update}
\begin{algorithm}[H]
\caption{Adversarial Policy Update}
\label{alg:adv}
\begin{algorithmic}[1]
\State Maintain strategy set $\{a^{(j)}\}_{j=1}^{J}$ with scores $R_j$.
\For{each minibatch}
  \State Sample $a^{(j)}$ with probability $\propto \exp(\beta R_j)$.
  \State Apply $a^{(j)}$; compute loss increase $\Delta \mathcal{J}$ and cost $\mathcal{C}_{\text{adv}}$.
  \State Update $R_j \leftarrow (1-\rho)R_j + \rho\,(\Delta \mathcal{J} - \lambda_{\text{ATP}}\mathcal{C}_{\text{adv}})$.
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Diagrams}
\subsection{Module Block Diagram}
\begin{figure}[H]
  \centering
  \begin{tikzpicture}[node distance=1.8cm,>=stealth,thick]
    \tikzstyle{blk}=[rectangle,rounded corners,draw,align=center,minimum width=3.4cm,minimum height=1.1cm]
    \node[blk] (ingest) {Evidence Ingest\\$\mathcal{E},\mathcal{G}$};
    \node[blk,right=2.6cm of ingest] (learn) {Mzekezeke\\Temporal Bayesian Learning};
    \node[blk,below=1.2cm of learn] (adv) {Diggiden\\Adversarial Hardening};
    \node[blk,right=2.8cm of learn] (orch) {Diadochi\\Orchestration/Ensembles};
    \node[blk,below=1.2cm of orch] (ctrl) {Hatata\\Decision Optimization};
    \node[blk,right=2.8cm of orch] (out) {Reconstruction\\$\hat{y}$ \, \\ Diagnostics};
    \draw[->] (ingest) -- (learn);
    \draw[->] (learn) -- (orch);
    \draw[->] (orch) -- (out);
    \draw[->] (learn) -- (adv);
    \draw[->] (adv) -- (learn);
    \draw[->] (orch) -- (ctrl);
    \draw[->] (ctrl) |- (orch);
  \end{tikzpicture}
  \caption{Module-level dataflow.}
\end{figure}

\subsection{Decay Function Families}
\begin{figure}[H]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=0.8\textwidth, height=6cm,
      xlabel={$\Delta t$}, ylabel={$\omega(\Delta t)$},
      legend style={at={(0.98,0.98)},anchor=north east},
      ymin=0, ymax=1.05
    ]
      \addplot[blue,thick] table[row sep=crcr] {%% dummy samples
0 1
1 0.61
2 0.37
3 0.22
4 0.14
5 0.08
}; \addlegendentry{Exponential}
      \addplot[red,thick,dashed] table[row sep=crcr] {%% power law-like
0 1
1 0.67
2 0.5
3 0.4
4 0.33
5 0.29
}; \addlegendentry{Power law}
      \addplot[green!60!black,thick,dash dot] table[row sep=crcr] {%% logistic-like
0 0.98
1 0.95
2 0.88
3 0.73
4 0.5
5 0.27
}; \addlegendentry{Logistic}
    \end{axis}
  \end{tikzpicture}
  \caption{Representative decay weight families.}
\end{figure}

\section{Metrics and Protocols}
\subsection{Reconstruction Consistency}
\begin{definition}[Consistency Score]
Let $\mathcal{C}\_\text{logical}$ be a set of constraints and $\mathcal{C}\_\text{emp}$ empirical checks. The consistency score is
\begin{equation}
 S_{\text{cons}} = \alpha \cdot \frac{|\{c \in \mathcal{C}\_\text{logical} : c\text{ satisfied}\}|}{|\mathcal{C}\_\text{logical}|} + (1-\alpha) \cdot \frac{|\{c \in \mathcal{C}\_\text{emp} : c\text{ passed}\}|}{|\mathcal{C}\_\text{emp}|}.
\end{equation}
\end{definition}

\subsection{Calibration}
Expected Calibration Error (ECE) on probabilistic outputs $\{(p_j, y_j)\}$ with $M$ bins:
\begin{equation}
 \mathrm{ECE} = \sum_{m=1}^{M} \frac{|B_m|}{n} \Bigl|\, \mathrm{acc}(B_m) - \mathrm{conf}(B_m)\,\Bigr|.
\end{equation}

\subsection{Robustness Under Attacks}
For a distribution over attacks $\Pi$, define relative degradation
\begin{equation}
 D_{\text{rob}} = \mathbb{E}_{a\sim \Pi}\Bigl[ \ell\bigl(\hat{y}; a(\mathcal{E},\mathcal{G})\bigr) - \ell\bigl(\hat{y}; (\mathcal{E},\mathcal{G})\bigr) \Bigr],
\end{equation}
with $\ell$ a task loss. Lower is better.

\subsection{Resource Accounting}
Throughput per unit cost:
\begin{equation}
 T_{\text{ATP}} = \frac{\text{processed\ items}}{\mathcal{C}_{\text{total}}}, \qquad \text{Latency--cost frontier: } \min\limits_{\text{configs}} (\text{latency}, \mathcal{C}_{\text{total}}).
\end{equation}

\section{Convergence and Stability}
\begin{theorem}[Existence of Minimizers]
Assume: (i) $q_{\Phi}$ belongs to a tight variational family; (ii) likelihoods in \eqref{eq:generative} are coercive in $\mathbf{z}$ under weights $\omega_i$; (iii) $\mathcal{C}_{\text{learn}}$ is lower semicontinuous. Then minimizers of \eqref{eq:relbo} exist.
\end{theorem}
\begin{proof}[Sketch]
Coercivity and tightness yield compact sublevel sets; lower semicontinuity implies existence by Weierstrass.
\end{proof}

\begin{proposition}[Robust Minimax Stationarity]
If $\mathcal{A}$ is compact in the attack topology and $\mathcal{J}_{\text{learn}}$ is convex in $(\Phi,\Theta)$ and upper semicontinuous in $a$, then a saddle point for \eqref{eq:minimax} exists.
\end{proposition}
\begin{proof}[Sketch]
Apply Sion's minimax theorem under stated regularity.
\end{proof}

\section{Implementation Notes}
The reference implementation organizes these modules as Rust crates with explicit resource accounting and orchestration layers. Expert adapters expose uniform interfaces; the orchestration module selects pattern $\kappa(c)$ per \eqref{eq:selector}. Cost ledgers are recorded for each operation to compute \eqref{eq:totalcost}. Repository structure and components are documented in the upstream source.

\section*{Acknowledgments}
We acknowledge prior architectural descriptions and repository organization that informed this specification.

\begin{thebibliography}{9}
\bibitem{honjo-repo}
Honjo Masamune repository. URL: \url{https://github.com/fullscreen-triangle/honjo-masamune}
\end{thebibliography}

\end{document}
